### 2018-06-17 paper/5-implementations.tex
Finish S5.3

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = pure (getValue k store)
            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-4mm}
\end{figure}

\vspace{-2mm}
\subsection{\Shake}\label{sec-implementation-shake}
\vspace{-1mm}

The \Shake approach for dependency tracking involves recording traces and
verifying them, for which we use the \hs{Trace} type defined
in~\ref{sec-verifying-traces}. Complete code is given
in~\S\ref{fig-shake-implementation}, split into three functions:

\begin{description}
\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the
    \hs{key} you are currently building, and a function \hs{check} which checks
    a specified dependency. It returns the \hs{result} field of all traces that
    match. Since \hs{check} is in an arbitrary monad, the function has to use
    \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&}
    functions\footnote{These functions are all available in the \cmd{extra}
    library on Hackage.}.
\item[\hs{recursive}] defines the dependency ordering pattern. It requires a
    \hs{process} function that builds a key given a way to recursively build a
    dependency, and a way to run task to produce result. The main purpose of
    \hs{recursive} is to ensure that in a single run no key is built twice~--~for
    which it extends the \hs{State} monad with a list of \hs{done} keys.
\item[\hs{shake}] ties everything together. First it tests if the traces allow
the current state of the target key. If not, it builds the key and updates the
store. The only subtlety is that Shake calls \hs{fetch} on the keys while
checking them~--~building the last-recorded dependencies before checking them.
One minor annoyance is that the \hs{State} has been extended and thus needs to
be projected using \hs{fst} before it is used.
\end{description}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
suspending rebuilder tasks key store = fst $ execState (build key) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) v
    build key = case tasks key of
        Nothing -> gets (getValue key . fst)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do v <- lift (build k)
                                 put =<< lift (gets (getInfo . fst))
                                 return v
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask fetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
            gets (getValue key . fst)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    vt <- get
    dirty <- not <$> verifyVT key value (fmap hash . fetch) vt
    if not dirty
    then return value
    else do
        (newValue, deps) <- trackM task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{1mm}
\end{figure}

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = return (getValue k store)
            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

\vspace{-1mm}
\subsection{\Shake}\label{sec-implementation-shake}
\vspace{-1mm}

Our model of \Shake is shown in Fig.~\ref{fig-shake-implementation}. It stores
verifying traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces}
as persistent build information and is composed of the \hs{suspending} scheduler
and \hs{vtRebuilder}.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- build the key
                                 lift (gets (getInfo . fst)) >>= put -- save new traces
                                 lift (gets (getValue k . fst))      -- fetch the value
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask fetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

The rebuilder executes the verification query \hs{verifyVT} to determine if the
\hs{key} is \hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s
current \hs{value}. Otherwise it executes the \hs{task}, obtaining both a
\hs{newValue} and the \hs{key}'s dynamic dependencies \hs{deps} (see the
definition of \hs{track} in~\S\ref{sec-deps}), which are subsequently recorded
in a new verification trace using \hs{recordVT}.

The \hs{suspending} scheduler is based on a recursive function \hs{build} that
builds a given \hs{key}, making sure not to duplicate work when called on the
same \hs{key} again in future. To achieve that, it keeps track of keys that have
already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}, running the
computation in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad.

Given a non-input \hs{key} that has not yet been built, we use the supplied
\hs{rebuilder} to embed the build information~\hs{i} into the \hs{task}.
We then execute the obtained \hs{newTask} with the \hs{fetch} callback, which
performs the following three steps when given a dependency \hs{k}:

\begin{itemize}
    \item Build \hs{k} by recursively calling the \hs{build} function. Note
          that if \hs{k} is dirty, this will effectively suspend the build of
          the current \hs{key} until \hs{k} is brought up to date.
    \item Save any new traces that have been generated when building \hs{k}.
    \item Read the up to date value of \hs{k} from the store and return it.
\end{itemize}

These steps are run in the
\hs{StateT}~\hs{i}~\hs{(@@State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k))}
monad, because the \hs{newTask} has access only to the~\hs{i} part of the state;
we use \hs{lift} to execute commands in the inner monad.

Finally, the \hs{newValue} and \hs{newInfo} obtained by running the \hs{newTask}
are stored, and \hs{key} is added to the set \hs{done}.
